---
title: "Statistical Analysis Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Analysis Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ZinaSuite)
library(ggplot2)
```

## Overview

This vignette demonstrates the statistical analysis capabilities of ZinaSuite for cancer genomics data.

## Basic Statistical Functions

### Correlation Analysis

```{r}
# Create sample data for correlation analysis
set.seed(123)
n <- 100
gene1_expr <- rnorm(n, mean = 5, sd = 1)
gene2_expr <- gene1_expr * 0.6 + rnorm(n, mean = 0, sd = 0.8)

# Calculate correlation
cor_result <- cor.test(gene1_expr, gene2_expr, method = "pearson")
print(cor_result)

# Create correlation plot
cor_data <- data.frame(
  Gene1 = gene1_expr,
  Gene2 = gene2_expr
)

p_cor <- ggplot(cor_data, aes(x = Gene1, y = Gene2)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  annotate("text", x = min(gene1_expr), y = max(gene2_expr), 
           label = sprintf("r = %.3f, p = %.3f", cor_result$estimate, cor_result$p.value),
           hjust = 0, size = 5) +
  theme_zinasuite() +
  labs(
    title = "Gene Correlation Analysis",
    x = "Gene 1 Expression",
    y = "Gene 2 Expression"
  )

print(p_cor)
```

### Multiple Correlation Tests

```{r}
# Create correlation matrix for multiple genes
set.seed(123)
genes <- c("TP53", "BRCA1", "EGFR", "MYC", "KRAS")
n <- 50
expr_matrix <- matrix(rnorm(n * length(genes)), nrow = n)
colnames(expr_matrix) <- genes

# Calculate correlation matrix
cor_matrix <- cor(expr_matrix)
print(round(cor_matrix, 3))

# Visualize correlation matrix
cor_long <- data.frame(
  Gene1 = rep(genes, each = length(genes)),
  Gene2 = rep(genes, length(genes)),
  Correlation = as.vector(cor_matrix)
)

p_cor_matrix <- ggplot(cor_long, aes(x = Gene1, y = Gene2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_zinasuite() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Gene Expression Correlation Matrix")

print(p_cor_matrix)
```

## Group Comparison

### T-Test

```{r}
# Create sample data for two groups
set.seed(123)
group1 <- rnorm(50, mean = 5, sd = 1)
group2 <- rnorm(50, mean = 6.5, sd = 1.2)

# Perform t-test
t_test_result <- t.test(group1, group2)
print(t_test_result)

# Visualize
test_data <- data.frame(
  Expression = c(group1, group2),
  Group = rep(c("Control", "Treatment"), each = 50)
)

p_ttest <- ggplot(test_data, aes(x = Group, y = Expression, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Control" = "#56B4E9", "Treatment" = "#E69F00")) +
  annotate("text", x = 1.5, y = max(test_data$Expression),
           label = sprintf("p = %.4f", t_test_result$p.value), size = 5) +
  theme_zinasuite() +
  labs(title = "Group Comparison (T-Test)")

print(p_ttest)
```

### ANOVA

```{r}
# Create sample data for multiple groups
set.seed(123)
group_a <- rnorm(30, mean = 5, sd = 1)
group_b <- rnorm(30, mean = 6, sd = 1)
group_c <- rnorm(30, mean = 7, sd = 1)

anova_data <- data.frame(
  Value = c(group_a, group_b, group_c),
  Group = rep(c("A", "B", "C"), each = 30)
)

# Perform ANOVA
anova_result <- aov(Value ~ Group, data = anova_data)
summary(anova_result)

# Post-hoc test
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)

# Visualize
p_anova <- ggplot(anova_data, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = get_palette("nature", n = 3)) +
  theme_zinasuite() +
  labs(title = "ANOVA: Multiple Group Comparison")

print(p_anova)
```

## Survival Analysis

### Survival Data Preparation

```{r}
# Create sample survival data
set.seed(123)
n <- 200
surv_data <- data.frame(
  time = rexp(n, rate = 0.1),
  status = sample(0:1, n, replace = TRUE, prob = c(0.3, 0.7)),
  group = sample(c("High", "Low"), n, replace = TRUE)
)

# Summary statistics
summary(surv_data)
```

### Kaplan-Meier Estimation

```{r}
# Calculate survival probabilities by group
surv_by_group <- aggregate(time ~ group, data = surv_data, 
                           FUN = function(x) c(mean = mean(x), median = median(x)))
print(surv_by_group)

# Create survival curve visualization
p_surv <- ggplot(surv_data, aes(x = time, color = group)) +
  stat_ecdf(geom = "step", size = 1.2) +
  scale_color_manual(values = c("High" = "#E69F00", "Low" = "#56B4E9")) +
  theme_zinasuite() +
  labs(
    title = "Survival Curve by Group",
    x = "Time",
    y = "Survival Probability",
    color = "Risk Group"
  )

print(p_surv)
```

### Log-Rank Test

```{r}
# Compare survival between groups
high_risk <- surv_data$time[surv_data$group == "High"]
low_risk <- surv_data$time[surv_data$group == "Low"]

# Wilcoxon test as approximation
log_rank_approx <- wilcox.test(high_risk, low_risk)
print(log_rank_approx)
```

## Dimension Reduction

### Principal Component Analysis (PCA)

```{r}
# Create sample high-dimensional data
set.seed(123)
n_samples <- 100
n_genes <- 20
expr_data <- matrix(rnorm(n_samples * n_genes), nrow = n_samples)
colnames(expr_data) <- paste0("Gene", 1:n_genes)

# Add group structure
groups <- sample(c("Cancer", "Normal"), n_samples, replace = TRUE)
expr_data[groups == "Cancer", 1:10] <- expr_data[groups == "Cancer", 1:10] + 2

# Perform PCA
pca_result <- prcomp(expr_data, scale. = TRUE)

# Extract PC scores
pca_scores <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Group = groups
)

# Calculate variance explained
var_explained <- summary(pca_result)$importance[2, 1:2] * 100

# Visualize PCA
p_pca <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("Cancer" = "#E69F00", "Normal" = "#56B4E9")) +
  theme_zinasuite() +
  labs(
    title = "PCA of Gene Expression Data",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  )

print(p_pca)

# Scree plot
scree_data <- data.frame(
  PC = 1:10,
  Variance = summary(pca_result)$importance[2, 1:10] * 100
)

p_scree <- ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(aes(group = 1), color = "red", size = 1) +
  geom_point(color = "red", size = 3) +
  theme_zinasuite() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Variance Explained (%)"
  )

print(p_scree)
```

## Regression Analysis

### Linear Regression

```{r}
# Create sample data for regression
set.seed(123)
n <- 100
x <- rnorm(n, mean = 5, sd = 2)
y <- 2 + 0.5 * x + rnorm(n, mean = 0, sd = 1)

# Fit linear model
lm_model <- lm(y ~ x)
summary(lm_model)

# Create regression plot
reg_data <- data.frame(x = x, y = y)

p_reg <- ggplot(reg_data, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE, fill = "pink") +
  annotate("text", x = min(x), y = max(y),
           label = sprintf("y = %.2f + %.2fx\nR² = %.3f",
                          coef(lm_model)[1], coef(lm_model)[2],
                          summary(lm_model)$r.squared),
           hjust = 0, size = 4) +
  theme_zinasuite() +
  labs(
    title = "Linear Regression",
    x = "Predictor",
    y = "Response"
  )

print(p_reg)
```

### Multiple Regression

```{r}
# Create sample data with multiple predictors
set.seed(123)
n <- 100
x1 <- rnorm(n)
x2 <- rnorm(n)
y <- 1 + 0.5 * x1 - 0.3 * x2 + rnorm(n, sd = 0.5)

# Fit multiple regression
multi_lm <- lm(y ~ x1 + x2)
summary(multi_lm)

# Coefficient plot
coef_data <- data.frame(
  Variable = names(coef(multi_lm))[-1],
  Coefficient = coef(multi_lm)[-1],
  CI_lower = confint(multi_lm)[-1, 1],
  CI_upper = confint(multi_lm)[-1, 2]
)

p_coef <- ggplot(coef_data, aes(x = Variable, y = Coefficient)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_zinasuite() +
  labs(title = "Regression Coefficients with 95% CI")

print(p_coef)
```

## Statistical Tests Summary

### Test Selection Guide

```{r}
# Create a summary table of statistical tests
test_summary <- data.frame(
  Test = c("T-Test", "ANOVA", "Wilcoxon", "Kruskal-Wallis", 
           "Chi-Square", "Correlation", "Linear Regression"),
  Purpose = c("Compare 2 groups (parametric)",
              "Compare 3+ groups (parametric)",
              "Compare 2 groups (non-parametric)",
              "Compare 3+ groups (non-parametric)",
              "Test independence of categorical variables",
              "Test association between continuous variables",
              "Model relationship between variables"),
  Data_Type = c("Continuous", "Continuous", "Continuous/Rank",
                "Continuous/Rank", "Categorical", "Continuous", "Continuous")
)

print(test_summary)
```

## Power Analysis

### Sample Size Calculation

```{r}
# Calculate effect size (Cohen's d)
cohens_d <- function(x, y) {
  pooled_sd <- sqrt(((length(x) - 1) * var(x) + (length(y) - 1) * var(y)) / 
                    (length(x) + length(y) - 2))
  (mean(x) - mean(y)) / pooled_sd
}

# Example calculation
set.seed(123)
group1 <- rnorm(50, mean = 5, sd = 1)
group2 <- rnorm(50, mean = 6, sd = 1)

effect_size <- cohens_d(group1, group2)
cat(sprintf("Cohen's d = %.3f\n", effect_size))

# Interpret effect size
if (abs(effect_size) < 0.2) {
  interpretation <- "Small effect"
} else if (abs(effect_size) < 0.5) {
  interpretation <- "Small to medium effect"
} else if (abs(effect_size) < 0.8) {
  interpretation <- "Medium to large effect"
} else {
  interpretation <- "Large effect"
}
cat(sprintf("Interpretation: %s\n", interpretation))
```

## Best Practices

### 1. Check Assumptions

```{r}
# Normality test
shapiro_test <- shapiro.test(gene1_expr)
cat(sprintf("Shapiro-Wilk test p-value: %.4f\n", shapiro_test$p.value))

# Equal variance test (Levene's test approximation)
var_test <- var.test(group1, group2)
cat(sprintf("F-test for equal variances p-value: %.4f\n", var_test$p.value))
```

### 2. Handle Multiple Testing

```{r}
# Example: Multiple correlation tests
genes <- paste0("Gene", 1:10)
p_values <- runif(10, 0.001, 0.1)  # Simulated p-values

# Bonferroni correction
p_adjusted_bonf <- p.adjust(p_values, method = "bonferroni")

# FDR correction (Benjamini-Hochberg)
p_adjusted_fdr <- p.adjust(p_values, method = "fdr")

# Summary table
mult_test_summary <- data.frame(
  Gene = genes,
  Raw_P = round(p_values, 4),
  Bonferroni = round(p_adjusted_bonf, 4),
  FDR = round(p_adjusted_fdr, 4)
)

print(mult_test_summary)
```

### 3. Report Effect Sizes

```{r}
# Calculate and report effect sizes
report_effect_size <- function(group1, group2) {
  d <- cohens_d(group1, group2)
  r <- cor(group1, group2)
  
  list(
    cohens_d = d,
    correlation = r,
    interpretation = ifelse(abs(d) < 0.2, "Small",
                           ifelse(abs(d) < 0.5, "Small-Medium",
                                  ifelse(abs(d) < 0.8, "Medium-Large", "Large")))
  )
}

# Example usage
effect_report <- report_effect_size(group1, group2)
print(effect_report)
```

## Session Information

```{r}
sessionInfo()
```

## Learn More

- `vignette("introduction")` - Package overview
- `vignette("data-query")` - Data querying guide
- `vignette("visualization")` - Visualization techniques
- [R for Data Science](https://r4ds.had.co.nz/)
- [The R Inferno](https://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
